Counterfactual Training: Teaching Models Plausible and Actionable Explanations

### 2025-03-15 paper/preprint/paper.tex

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Counterfactual Training (A Preprint) }
\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Patrick
Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
Engineering, Mathematics and Computer Science\\Delft University of
Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Aleksander
Buszydlik}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Arie van
Deursen}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Cynthia C. S.
Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\}
\date{}
\begin{document}
\maketitle
\begin{abstract}
We propose a novel training regime termed counterfactual training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be (1) plausible with
respect to the underlying data and (2) actionable with respect to the
user-defined mutability constraints. Much existing research has
therefore focused on developing post-hoc methods to generate
counterfactuals that meet these desiderata. In this work, we instead
hold models directly accountable for the desired end goal:
counterfactual training employs counterfactuals ad-hoc during the
training phase to minimize the divergence between learned
representations and plausible, actionable explanations. We demonstrate
empirically and theoretically that our proposed method facilitates
training models that deliver inherently desirable explanations while
maintaining high predictive performance.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Training \sep Counterfactual
Explanations \sep Algorithmic Recourse \sep Explainable AI \sep 
Representation Learning

@@

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI), a paradigm that focuses on the
development of tools to derive (post-hoc) explanations from complex
model representations. Such explanations should mitigate a scenario in
which practitioners deploy opaque models and blindly rely on their
predictions. On countless occasions, this has happened in practice and
caused real harms to people who were adversely and unfairly affected by
automated decision-making (ADM) systems involving opaque models
(\citeproc{ref-mcgregor2021preventing}{McGregor 2021};
\citeproc{ref-oneil2016weapons}{O'Neil 2016}). Effective XAI tools can
aid us in monitoring models and providing recourse to individuals to
turn negative outcomes (e.g., ``loan application rejected'') into
positive ones (e.g., ``application accepted''). In line with this, our
work builds upon \textbf{counterfactual explanations} (CE) proposed by
Wachter, Mittelstadt, and Russell
(\citeproc{ref-wachter2017counterfactual}{2017}) as an effective
approach to achieve this goal. CEs prescribe minimal changes for factual
inputs that, if implemented, would prompt some fitted model to produce a
desired output.

@@

This interpretation of the link between explainability through
counterfactuals on one side and robustness to adversarial examples on
the other is backed by empirical evidence. Sauer and Geiger
(\citeproc{ref-sauer2021counterfactual}{2021}) demonstrates that using
counterfactual images during classifier training improves model
robustness. Similarly, Abbasnejad et al.
(\citeproc{ref-abbasnejad2020counterfactual}{2020}) argue that
counterfactuals represent potentially useful training data in machine
learning, especially in supervised settings where inputs may be
reasonably mapped to multiple outputs. They, too, demonstrate that
augmenting the training data of image classifiers can improve
generalization. Finally, Teney, Abbasnedjad, and Hengel
(\citeproc{ref-teney2020learning}{2020}) propose an approach using
counterfactuals in training that does not rely on data augmentation:
they argue that counterfactual pairs typically already exist in training
datasets. Specifically, their approach relies on identifying similar
input samples with different annotations and ensuring that the gradient
of the classifier aligns with the vector between such pairs of
counterfactual inputs using the cosine distance as the loss function.

@@

Our key outcome of interest is improved explainability (Def.
\ref{def-explainability}). To this end, we focus primarily on the
plausibility and cost of faithfully generated counterfactuals at test
time. To measure the cost of counterfactuals, we follow the standard
convention of using distances (\(\ell_1\)-norm) between factuals and
counterfactuals as a proxy. For plausibility, we assess how similar
counterfactuals are to observed samples in the target domain,
\(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the distance-based
metric used in Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}),

@@

To assess our proposed training regime, we investigate the improvements
in performance metrics when using CT on top of a weak baseline (BL):
specifically, a multilayer perceptron (\emph{MLP}). This is the best way
to get a clear picture of how effective CT is and consistent with how
assessment is done in the related literature
(\citeproc{ref-ross2021learning}{Ross, Lakkaraju, and Bastani 2024};
\citeproc{ref-teney2020learning}{Teney, Abbasnedjad, and Hengel 2020};
\citeproc{ref-goodfellow2014explaining}{Goodfellow, Shlens, and Szegedy
2015}).

@@

Table~\ref{tbl-main} presents our main empirical findings. For all
datasets except \emph{OL} and across all test settings, the average
distance of CEs from observed samples in the target class is reduced,
indicating improved plausibility. The magnitude of improvements varies.
For the simple synthetic datasets, distance reductions range from around
20-40\% (\emph{LS}, \emph{Moon}) to almost 60\% (\emph{Circ}). For the
real-world tabular datasets, improvements tend to be smaller but still
substantial, with around 10-15\% for \emph{CH}, 11-28\% for \emph{GMSC},
7-8\% for \emph{Cred}, and around 3\% for \emph{Adult}. For our only
vision dataset (\emph{MNIST}), distances are reduced by up to 9\%. The
results for our proposed divergence metric are qualitatively similar,
but generally even more pronounced: for the \emph{Circ} dataset,
implausibility is reduced by almost 94\% to virtually zero as we
verified by the absolute outcome. Improvements for other datasets range
from 28\% (\emph{Moon}) to 78\% (\emph{GMSC}). For \emph{OL} the
reduction is negative, consistent with the distance-based metric.
\emph{MNIST} is the only dataset for which the distance and divergence
metrics disagree. Upon visual inspection of the image counterfactuals we
find that CT clearly improves plausibility (see appendix).

@@

\section{Discussion}\label{sec-discussion}

As our results indicate, counterfactual training produces models that
are more explainable. Nonetheless, it brings about three important
limitations.\\
\indent \emph{CT increases the training time of models.} CT can be more
time-consuming than conventional training regimes. While higher numbers
of CEs per iteration positively impact the quality of solutions, they
also increase the amount of computations. Relatively small grids with
270 settings can take almost four hours for more demanding datasets on a
high-performance computing cluster with 34 2GB CPUs.\footnote{See
  supplementary appendix for computational details.} Three factors
attenuate this effect. First, CT amortizes the cost of CEs for the
training samples. Second, we find that it can retain its value when used
as a ``fine-tuning'' technique for conventionally-trained models. Third,
it yields itself to parallel execution, which we have leveraged for our
own experiments.\\
\indent \emph{Immutable features may have proxies.} We propose an
approach to protect immutable features and thus increase the
actionability of the generated CEs. However, it requires that model
owners define the mutability constraints for (all) features considered
by the model. Even if all immutable features are protected, there may
exist proxies that are mutable (and hence should not be protected) but
preserve enough information about the principals to hinder the
protections. Delineating actionability is a major undecided challenge in
the AR literature (see, e.g.,
(\citeproc{ref-venkatasubramanian2020philosophical}{Venkatasubramanian
and Alfano 2020})) impacting the capacity of CT to fulfill its intended
goal.\\
\indent \emph{Interventions on features may impact fairness.} We provide
a tool that allows practitioners to modify the sensitivity of a model
with respect to certain features, which may have implication for the
fair and equitable treatment of decision subjects. As protecting a set
of features leads the model to assign higher relative importance to
unprotected features, model owners could misuse our solution by
enforcing explanations based on features that are more difficult to
modify by some (group of) individuals. For example, consider the Adult
dataset used in our experiments, where \emph{workclass} or
\emph{education} may be more difficult to change for underpriviledged
groups. When applied irresponsibly, CT could result in an unfairly
assigned burden of recourse (\citeproc{ref-sharma2020certifai}{Sharma,
Henderson, and Ghosh 2020}), threatening the equality of opportunity in
the system (\citeproc{ref-bell2024fairness}{Bell et al. 2024}).
Nonetheless, these phenomena are not specific to CT.

We also highlight several important directions for future research.
Firstly, it is an interesting challenge to extend CT beyond
classification settings. Our formulation relies on the distinction
between non-target class(es) \(y^{-}\) and target class(es) \(y^{+}\) to
generate counterfactuals through Equation~\ref{eq-obj}. While \(y^{-}\)
and \(y^{+}\) can be arbitrarily defined, CT requires the output space
\(\mathcal{Y}\) to be discrete. Thus, it does not apply to ML tasks
where the change in outcome cannot be readily quantified. Focus on
classification models is a common restriction in research on CEs and AR.
Other settings have attracted some interest (e.g., regression in
(\citeproc{ref-spooner2021counterfactual}{Spooner et al. 2021})), but
there is little consensus how to robustly extend the notion of CEs.\\
\indent Secondly, our approach is susceptible to training instabilities.
This problem has been recognized for JEMs
(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) and even
though we depart from the SGLD-based sampling, we still encounter
considerable variability in the outcomes. CT is exposed to two potential
sources of instabilities: (1) the energy-based contrastive divergence
term in Equation~\ref{eq-div}, and (2) the underlying counterfactual
explainers. We find several promising ways to mitigate this problem:
regularizing energy (\(\lambda_{\text{reg}}\)), generating sufficiently
many counterfactuals during each epoch, and including only mature
counterfactuals for contrastive divergence.\\
\indent Finally, we believe that it is possible to substantially improve
hyperparameter selection procedures. Our method benefits from the tuning
of certain key hyperparameters (see Section~\ref{sec-hyperparameters}).
In this work, we have relied exclusively on grid search for this task.
Future work on CT could benefit from investigating more sophisticated
approaches. Notably, CT is iterative which makes methods such as
Bayesian or gradient-based optimization applicable (see, e.g.,
(\citeproc{ref-bischl2023hyperparameter}{Bischl et al. 2023})).

\section{Conclusion}\label{sec-conclusion}

State-of-the-art machine learning models are prone to learning complex
representations that cannot be interpreted by humans and existing
post-hoc explainability approaches cannot guarantee that the
explanations agree with the model's learned representation of data. As a
step towards addressing this challenge, we introduced counterfactual
training, a novel training regime that incentivizes highly-explainable
models. Our approach leads to explanations that are both
plausible---compliant with the underlying data-generating process---and
actionable---compliant with user-specified mutability constraints---and
thus meaningful to their recipients. Through extensive experiments we
demonstrate that CT satisfies its objectives while preserving the
predictive performance of the models. Our approach can also be used to
fine-tune conventionally-trained models and achieve similar gains in
explainability. Finally, this work showcases that it is practical to
improve models \emph{and} their explanations at the same time.

@@

\bibitem[\citeproctext]{ref-mcgregor2021preventing}
McGregor, Sean. 2021. {``{Preventing repeated real world AI failures by
cataloging incidents: The AI incident database}.''} In \emph{Proceedings
of the AAAI Conference on Artificial Intelligence}, 35:15458--63. 17.

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-train-default}: Training Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-eval-default}: Evaluation Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-train-final-run-train}: Training Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-warning-color!10!white, colframe=quarto-callout-warning-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Package Version (Reproducibility)}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-tune-train}: Training Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-tune_lr-train}: Training Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]

>>>>

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Counterfactual Training (A Preprint) }
\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Patrick
Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
Engineering, Mathematics and Computer Science\\Delft University of
Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Aleksander
Buszydlik}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Arie van
Deursen}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Cynthia C. S.
Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\}
\date{}
\begin{document}
\maketitle
\begin{abstract}
We propose a novel training regime termed counterfactual training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be (1) plausible with
respect to the underlying data and (2) actionable with respect to the
user-defined mutability constraints. Much existing research has
therefore focused on developing post-hoc methods to generate
counterfactuals that meet these desiderata. In this work, we instead
hold models directly accountable for the desired end goal:
counterfactual training employs counterfactuals ad-hoc during the
training phase to minimize the divergence between learned
representations and plausible, actionable explanations. We demonstrate
empirically and theoretically that our proposed method facilitates
training models that deliver inherently desirable explanations while
promoting robustness and preserving high predictive performance.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Training \sep Counterfactual
Explanations \sep Algorithmic Recourse \sep Explainable AI \sep 
Representation Learning

@@

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI), a paradigm that focuses on the
development of tools to derive (post-hoc) explanations from complex
model representations. Such explanations should mitigate a scenario in
which practitioners deploy opaque models and blindly rely on their
predictions. On countless occasions, this has happened in practice and
caused real harms to people who were adversely and unfairly affected by
automated decision-making (ADM) systems involving opaque models (see,
e.g., (\citeproc{ref-oneil2016weapons}{O'Neil 2016})). Effective XAI
tools can aid us in monitoring models and providing recourse to
individuals to turn negative outcomes (e.g., ``loan application
rejected'') into positive ones (e.g., ``application accepted''). In line
with this, our work builds upon \textbf{counterfactual explanations}
(CE) proposed by Wachter, Mittelstadt, and Russell
(\citeproc{ref-wachter2017counterfactual}{2017}) as an effective
approach to achieve this goal. CEs prescribe minimal changes for factual
inputs that, if implemented, would prompt some fitted model to produce a
desired output.

@@

This interpretation of the link between explainability through
counterfactuals on one side and robustness to adversarial examples on
the other is backed by empirical evidence. Sauer and Geiger
(\citeproc{ref-sauer2021counterfactual}{2021}) demonstrates that using
counterfactual images during classifier training improves model
robustness. Similarly, Abbasnejad et al.
(\citeproc{ref-abbasnejad2020counterfactual}{2020}) argues that
counterfactuals represent potentially useful training data in machine
learning, especially in supervised settings where inputs may be
reasonably mapped to multiple outputs. They, too, demonstrate that
augmenting the training data of image classifiers can improve
generalization. Finally, Teney, Abbasnedjad, and Hengel
(\citeproc{ref-teney2020learning}{2020}) proposes an approach using
counterfactuals in training that does not rely on data augmentation:
they argue that counterfactual pairs typically already exist in training
datasets. Specifically, their approach relies on identifying similar
input samples with different annotations and ensuring that the gradient
of the classifier aligns with the vector between such pairs of
counterfactual inputs using the cosine distance as the loss function.

@@

Our key outcome of interest is improvement in explainability (Def.
\ref{def-explainability}). To this end, we focus primarily on the
plausibility and cost of faithfully generated counterfactuals at test
time. To measure the cost, we follow the standard convention of using
distances (\(\ell_1\)-norm) between factuals and counterfactuals as a
proxy. For plausibility, we assess how similar CEs are to the observed
samples in the target domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We
rely on the distance-based metric used in Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}),

@@

To assess CT, we investigate the improvements in performance metrics
when using it on top of a weak baseline (BL): a multilayer perceptron
(\emph{MLP}). This is the best way to get a clear picture of the
effectiveness of CT, and it is consistent with how assessment is done in
the related literature
(\citeproc{ref-goodfellow2014explaining}{Goodfellow, Shlens, and Szegedy
2015}; \citeproc{ref-ross2021learning}{Ross, Lakkaraju, and Bastani
2024}; \citeproc{ref-teney2020learning}{Teney, Abbasnedjad, and Hengel
2020}).

@@

Table~\ref{tbl-main} presents our main empirical findings. For all
datasets except \emph{OL} and across all test settings, the average
distance of CEs from observed samples in the target class is reduced,
indicating improved plausibility. The magnitude of improvements varies.
For the simple synthetic datasets, distance reductions range from around
20-40\% (\emph{LS}, \emph{Moon}) to almost 60\% (\emph{Circ}). For the
real-world tabular datasets, improvements tend to be smaller but still
substantial, with around 10-15\% for \emph{CH}, 11-28\% for \emph{GMSC},
7-8\% for \emph{Cred}, and around 3\% for \emph{Adult}. For the vision
dataset (\emph{MNIST}), distances are reduced by up to 9\%. The results
for our proposed divergence metric are qualitatively similar, but
generally even more pronounced: for the \emph{Circ} dataset,
implausibility is reduced by almost 94\% to virtually zero as we
verified by the absolute outcome. Improvements for other datasets range
from 28\% (\emph{Moon}) up to 78\% (\emph{GMSC}). For \emph{OL} the
reduction is negative, consistent with the distance-based metric.
\emph{MNIST} is the only dataset for which the distance and divergence
metrics disagree. Upon visual inspection of the image counterfactuals we
find that CT clearly improves plausibility (see supplementary appendix
for images).

@@

\section{Conclusions}\label{sec-discussion}

As our results indicate, counterfactual training produces models that
are more explainable. Nonetheless, it brings about three important
limitations.\\
\indent \emph{CT increases the training time of models.} CT can be more
time-consuming than conventional training regimes. While higher numbers
of CEs per iteration positively impact the quality of solutions, they
also increase the amount of computations. Relatively small grids with
270 settings can take almost four hours for more demanding datasets on a
high-performance computing cluster with 34 2GB CPUs.\footnote{See
  supplementary appendix for computational details.} Three factors
attenuate this effect: (1) CT amortizes the cost of CEs for the training
samples; (2) it can retain its value when used as a ``fine-tuning''
technique for conventionally-trained models; and (3) it yields itself to
parallel execution, which we have leveraged for our own experiments.\\
\indent \emph{Immutable features may have proxies.} We propose an
approach to protect immutable features and thus increase the
actionability of the generated CEs. However, it requires that model
owners define the mutability constraints for (all) features considered
by the model. Even if all immutable features are protected, there may
exist proxies that are mutable (and hence should not be protected) but
preserve enough information about the principals to hinder the
protections. Delineating actionability is a major open challenge in the
AR literature (see, e.g.,
(\citeproc{ref-venkatasubramanian2020philosophical}{Venkatasubramanian
and Alfano 2020})) impacting the capacity of CT to fulfill its intended
goal.\\
\indent \emph{Interventions on features may impact fairness.} We provide
a tool that allows practitioners to modify the sensitivity of a model
with respect to certain features, which may have implication for the
fair and equitable treatment of decision subjects. Model owners could
misuse this solution by enforcing explanations based on features that
are more difficult to modify by some (group of) individuals. For
example, consider the \emph{Adult} dataset used in our experiments,
where \emph{workclass} or \emph{education} may be more difficult to
change for underpriviledged groups. When applied irresponsibly, CT could
result in an unfairly assigned burden of recourse
(\citeproc{ref-sharma2020certifai}{Sharma, Henderson, and Ghosh 2020}),
threatening the equality of opportunity in the system
(\citeproc{ref-bell2024fairness}{Bell et al. 2024}). Nonetheless, these
phenomena are not specific to CT.\\
\hfill \break \indent We also highlight several important directions for
future research. Firstly, it is an interesting challenge to extend CT
beyond classification settings. Our formulation relies on the
distinction between non-target class(es) \(y^{-}\) and target class(es)
\(y^{+}\) to generate counterfactuals through Equation~\ref{eq-obj}.
While \(y^{-}\) and \(y^{+}\) can be arbitrarily defined, CT requires
the output space \(\mathcal{Y}\) to be discrete. Thus, it does not apply
to ML tasks where the change in outcome cannot be readily quantified.
Focus on classification models is a common restriction in research on
CEs and AR. Other settings have attracted some interest (e.g.,
regression in (\citeproc{ref-spooner2021counterfactual}{Spooner et al.
2021})), but there is little consensus how to robustly extend the notion
of CEs.\\
\indent Secondly, our approach is susceptible to training instabilities.
This problem has been recognized for JEMs
(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) and even
though we depart from the SGLD-based sampling, we still encounter
considerable variability in the outcomes. CT is exposed to two potential
sources of instabilities: (1) the energy-based contrastive divergence
term in Equation~\ref{eq-div}, and (2) the underlying counterfactual
explainers. We find several promising ways to mitigate this problem:
regularizing energy (\(\lambda_{\text{reg}}\)), generating sufficiently
many counterfactuals during each epoch, and including only mature
counterfactuals for contrastive divergence.\\
\indent Finally, we believe that it is possible to substantially improve
hyperparameter selection procedures. Our method benefits from the tuning
of certain key hyperparameters (see Section~\ref{sec-hyperparameters}).
In this work, we have relied exclusively on grid search for this task.
Future work on CT could benefit from investigating more sophisticated
approaches. Notably, CT is iterative which makes methods such as
Bayesian or gradient-based optimization applicable (see, e.g.,
(\citeproc{ref-bischl2023hyperparameter}{Bischl et al. 2023})).
\hfill \break

To conclude, state-of-the-art machine learning models are prone to
learning complex representations that cannot be interpreted by humans
and existing post-hoc explainability approaches cannot guarantee that
the explanations agree with the model's learned representation of data.
As a step towards addressing this challenge, we introduced
counterfactual training, a novel training regime that incentivizes
highly-explainable models. Our approach leads to explanations that are
both plausible---compliant with the underlying data-generating
process---and actionable---compliant with user-specified mutability
constraints---and thus meaningful to their recipients. Through extensive
experiments we demonstrate that CT satisfies its objective while
promoting robustness and preserving the predictive performance of the
models. It can also be used to fine-tune conventionally-trained models
and achieve similar gains. Finally, this work showcases that it is
practical to improve models \emph{and} their explanations at the same
time.

@@

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-train-default}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-eval-default}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-train-final-run-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Package Version (Reproducibility)}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-warning-color!10!white, colframe=quarto-callout-warning-color-frame, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-tune-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-tune_lr-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
